---
title: "Solutions for UKDD Challenges"
author: "Linh Dang"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
editor: visual
---

```{r}
#| code-fold: true
#| message: false
#| warning: false

library(Biostrings)
library(dplyr)
rm(list = ls())
```

## Challenge 1

```{r}
#| message: false
#| warning: false

seqs <- rprojroot::find_rstudio_root_file() %>% 
  file.path("data/KGE_DNA_test.fasta") %>% 
  Biostrings::readDNAStringSet()

## write a function to call reverseComplement
get_reverseComplement <- function(input_DNAstring, start_idx=1, end_idx=2){
  ## input_DNAstring    --> DNAString object
  ## start_idx, end_idx --> integer within proper range: 1..length(seq)
  ## end_idx >= start_idx > 0
  ##---------------------------------------------------
  ## check data type
  #tmp_seq <- NA
  if(!"DNAString" %in% class(input_DNAstring)){
    stop("input_DNAstring has to be DNAString object")
  }
  ## check if index is integer
  if(is.na(start_idx) || is.na(end_idx)){
    stop("Index is not valid!")
    #tmp_seq <- 
  }
  
  ## check condition
  
  ## return value
  res <- input_DNAstring %>% 
    subseq(start = start_idx, end = end_idx) %>% 
    Biostrings::reverseComplement()
  return(res)
}
idx_start <- 2
idx_end <- 10
res <- get_reverseComplement(seqs[[1]], idx_start, idx_end)
print(sprintf("Original seq: %s", subseq(seqs[[1]], idx_start, idx_end)))
print(sprintf("Reversed complement seq: %s", res))
```

## Challenge 2

Assumption two reads already aligned (have the same length)

```{r}
#| message: false
#| warning: false

transition_lst <- c("AG", "GA", "CT", "TC")
transition_num  <- 0
transversion_num <- 0
S1 <- seqs[[1]] %>% as.character()
S2 <- seqs[[2]] %>% as.character()
for(i in 1:nchar(S1)){
  if(substr(S1,i,i) == substr(S2,i,i)){next}
  if(paste0(substr(S1,i,i), substr(S2, i, i)) %in% transition_lst){
    transition_num <- transition_num +1
  }else{
    transversion_num <- transversion_num +1
  }
}
R <- transition_num/transversion_num
print(sprintf("transition/transversion ratio is: %f", R))
```

## Challenge 3

```{r}
#| code-fold: true
#| message: false
#| warning: false

t <- "ACGTTGCA"
S <- seqs[[1]] %>% as.character()
idxs <- Biostrings::gregexpr2(t, S)[[1]]
print(t)
## double check
for(i in idxs){
  print(sprintf("Sub-string starts with an index: %i --> %s", i, substr(seqs[[1]], i, (i+nchar(t)-1))))
}
```

## Challenge 4

Find maximal repeat of a string

```{r}
#| message: false
#| warning: false

## debug(Check_maximal_repeat)
## Check_maximal_repeat("ACGTAAATCGTAT", "CG")
Check_maximal_repeat <- function(S, t){
  t_idx <- gregexpr2(pattern = t, text = S)[[1]]
  if(length(t_idx)<=1){
    return(FALSE)
  }
  if(!1 %in% t_idx){
    ## expand to the left
    t_left_expand <- substr(S, t_idx[1]-1, t_idx[1]-1 + nchar(t))
    tmp_idx <- gregexpr2(t_left_expand, S)[[1]]
    tmp_idx <- tmp_idx +1
    if(length(setdiff(t_idx, tmp_idx))==0){
      return(FALSE)
    }
  }
  ## can only expand to the right
  t_right_expand <- substr(S, t_idx[1], t_idx[1] + nchar(t))
  tmp_idx <- gregexpr2(t_right_expand, S)[[1]]
  if(length(setdiff(t_idx, tmp_idx))==0){
    return(FALSE)
  }else{return(TRUE)}
}
S <- seqs[[1]] %>% as.character()
#S <- "TAGTTAGCGAGA"
min_bp <- 20
#t_init <- subseq(Sm 1, min_bp)

lst_maximal <- c()
num <- 0
## start from 1, go through the string
for(i in 1:(nchar(S) - min_bp+1)){
  for(j in (i+min_bp-1):nchar(S)){
    t <- subseq(S, i, j)
    if(length(gregexpr2(t,S)[[1]])==1){
      #print(sprintf("Break at (i,j) = (%d,%d)", i, j))
      break
      }
    if(Check_maximal_repeat(S,t) && !t %in% lst_maximal){
      lst_maximal <- c(lst_maximal, t)
      num <- num +1
      ##print(sprintf("%d: %s", num, t))
    }
  }
  if(num>=30){break}
}
print(lst_maximal)
```

## Challenge 5

```{r}
#| message: false
#| warning: false

library(fastqcr)
df <- rprojroot::find_rstudio_root_file() %>% 
  file.path("test_data/AN215152_S2_R1_001_fastqc/fastqc_data.txt") %>% 
  fastqcr::qc_read(modules = "Per sequence quality scores", verbose = F)
df <- df$per_sequence_quality_scores
## sum the number of reads whose quality score less than 30
res <- df %>% 
  filter(Quality <30) %>% 
  pull(Count) %>% 
  sum()
print(sprintf("Number of reads whose quality score is less than 30: %d", res))
```

## Challenge 6

### Observation

1\) If a set X has N elements, its corresponding difference multiset $\Delta(X)$ has N\*(N-1)/2 elements.

2\) Set X and set X + a (plus elementwise) have the same difference multiset.

3\) Suppose X has N elements and in order $X =\{x_1, x_2,...,x_N | x_1 < x_2 <...< x_N\}$, the biggest element in $\Delta(X)$ has to be $x_N - x_1$.

4\) The following $N-1$ elements could be used to determined the rest: $$\{(x_N - x_1), (x_N -x_2), ..., (x_N -x_{N-1})\}$$

### Approach

1\) Construct a linear system $Ax=B$ to capture the differences.

$$\begin{aligned}
x_N+0*x_{N-1}+...+0*x_2 -x_1 &= b_1 \\
x_N+0*x_{N-1}+...-x_2 +0*x_1 &= b_2 \\
... \\
x_N - x_{N-1}+...+0*x_2 + 0*x_1 &= b_{N-1}
\end{aligned}$$

where $b_1$ is the biggest element in $\Delta(X)$, and $b_2 > b_3 > ... > b_{N-1}$.

2\) The other differences are the difference among above linear equations. Thus the set \$\\{b_1, b_2,...,b\_{N-1}\\}\$ must have the difference multiset as \$\\{b_N,b\_{N+1},...\\}\$

3\) $X \rightarrow \Delta(X)$ is much easier to calculate than $X \leftarrow \Delta(X)$

```{r}
#| message: false
#| warning: false

library(pracma)
rm(list = ls())

## generate difference multiset of a given set
Diff_Multiset_generate <- function(X_input){
   X <- X_input[!duplicated(X_input)]
   if(length(X) != length(X_input)){
     print("coercing into a set")
   }
   N <- length(X)
   L <- c()
   for(i in 1:(N-1)){
     for (j in (i+1):N) {
       L <- c(L, abs(X[i]-X[j]))
     }
   }
   L <- sort(L)
   return(L)
}
##---------
## estimate original set from its corresponding difference multiset
Estimate_orignal <- function(L, max_iter_num = 100){
  ## L is multiset containing positive elements
  L <- sort(L, decreasing = T)
  N <- (1 + sqrt(8*length(L)+1))/2
  if(N != round(N)){
    stop("Number of elements in difference delta multiset is wrong!!!")
  }
  ## suppose that N >=3, in this case, when dealing with A*x = B, we will have the number of columns is greater than number of rows. Thus we muss apply decomposition to solve the matrix system
  ## A has dimension (N-1) x N
  A <- zeros(n=N-1, m = N)
  A[,N] <- 1
  A[,1:(N-1)] <- -1*diag(N-1)
  ## It is more tricky to construct B. We know that the first value of B has to be biggest value in L, but the rest are not sure, but they have to be distinct. 
  ## One way to tackle is that we compute all possibilites, and check the validation, as mentioned aboved.
  B <- c(L[1])
  L_reduced <- L[2:length(L)]
  L_reduced <- L_reduced[!duplicated(L_reduced)]
  All_cmbn <- combn(L_reduced, N-2)
  Solutions <- c()
  iter_num <- 0
  ##max_iter_num <- 100
  for(i in 1:ncol(All_cmbn)){
    if(iter_num > max_iter_num){
      ## the number of all combinations could be huge, thus only check with certain amount. 
      print(sprintf("Maximum number of iteration: %d has been reached! Increase if needed", max_iter_num))
      break}
    B <- c(L[1])
    B <- c(B, All_cmbn[,i]) %>% sort(decreasing = T)
    rest <- L[-match(B,L)] %>% sort(decreasing = T)
    B_delta <- Diff_Multiset_generate(B) %>% sort(decreasing = T)
    if(!all(rest == B_delta)){
      ## wrong guest, move to next combination
      iter_num <- iter_num +1
      next
    }
    asvd = svd(A)
    adiag = diag(1/asvd$d)
    #adiag <- adiag *1
    solution = asvd$v %*% adiag %*% t(asvd$u) %*% B
    ## shift solution
    min_val <- min(solution)
    solution <- solution + (1-min_val)
    Solutions <- cbind(Solutions, solution)
    ##print(Solutions)
    iter_num <- iter_num +1
    ##
  }
  return(Solutions)
}
####-------------------

#L <- c(2,2,3,3,4,5,6,7,8,10)
X <- c(1,4,9,14,19,20,25)
L <- Diff_Multiset_generate(X)
Solutions <- Estimate_orignal(L, max_iter_num = 1000)
print(Solutions)
```

**Updated version**

Using prior knowledge to get B instead of go through all random combinations. And note that if we shift the set to $-x_1$ ($x_1$ become 0 and $x_N$ is the biggest element in the difference multiset), we have the same difference multiset. Thus, the values of vector $B$ additional with $0$ is the set we want to find.

```{r}
#| message: false
#| warning: false

## calculate the upper bound of index to reduce computational cost.
## value from the lower-left rectangle
upper_bound <- function(i, N) {
  (i - 1) * N - i * (i - 1) / 2 + 1
}
# check if a vec is fitted with L
## check if delta(vec) is sub-multiset of L\vec
check_validity <- function(L, vec){
  #vec <- vec %>% sort(decreasing = T)
  #return(grepl(paste(Diff_Multiset_generate(vec) %>% sort(decreasing = T),collapse=";"),paste(L_reduced,collapse=";"))>0)
  A <- Diff_Multiset_generate(vec) %>% sort(decreasing = T)
  B <- L[-match(vec,L)]
  freq_A <- table(A)
  freq_B <- table(B)
  #print(freq_A)
  #print(freq_B)
  
  # All elements of A must exist in B with sufficient count
  all(
    sapply(names(freq_A), function(el) {
      (freq_B[el] %in% NA) == FALSE &&   # element exists in B
        freq_A[el] <= freq_B[el]            # B has at least as many copies
    })
  )
}
##----------------------------
pick_all_elements <- function(L) {
  
  N <- (1 + sqrt(8*length(L)+1))/2
  if(N != round(N)){
    stop("Number of elements in difference delta multiset is wrong!!!")
  }
  
  all_solutions <- list()
  # Recursive backtracking
  # i       : current pick (1 to N-1)
  # min_idx : to keep the last picked index
  backtrack <- function(i, min_idx, picked_vals) {
    
    if (i > N - 1){
      all_solutions[[length(all_solutions) + 1]] <<- picked_vals ## save picked_value to global var
      return(NULL)
    } ##return(list())  # All N-1 elements picked successfully
    
    ub <- upper_bound(i,N)           # Allowed range: L[1:ub], but must be > min_idx
    
    # Valid indices for this pick: must be in [1, ub] AND > min_idx
    valid_indices <- seq(min_idx + 1, ub)
    
    if (length(valid_indices) == 0) return(NULL)  # Dead end
    
    # Randomly shuffle to get a random valid solution (not always the first)
    for (idx in seq(min_idx + 1, ub)) {
      ## check 
      candidate_vals <- c(picked_vals, L[idx])
      if(length(candidate_vals) >1){
        ## check if two same elements in a set
        if(length(candidate_vals) != length(candidate_vals[!duplicated(candidate_vals)])) next
        ## check if a picked value satisfied the property of complete.
        if (!check_validity(L, candidate_vals)) next
        
      }
      ## next round
      backtrack(i + 1, idx, candidate_vals)
    }
  }
  backtrack(1, 0, c())
  
  if (length(all_solutions) == 0) {
    message("No valid selection found.")
    return(NULL)
  }
  
  Solutions <- all_solutions[!duplicated(as.list(all_solutions))]
  mx <- sapply(Solutions, function(x)cbind(x))
  return(rbind(mx, 0))
  
}
### Test ------------------
X <- c(1,4,9,14,19,20,25)
L <- Diff_Multiset_generate(X) %>% sort(decreasing = T)
Solutions <- pick_all_elements(L)
print(Solutions)
```

## Remarks

Overall, I have implemented those functions with expectations that users will give the proper inputs. The next step could be:

- Better: Check and restrict users' inputs and announce errors whenever encountering wrong inputs.

- Best: Implementing unit testing for each functions.

The follows are the remarks for each challenge:

### Challenge 1

At the moment the function only except input as a sequence of DNAString object, but it could be easily adjusted to accept string of character.

Start and end indices when sub-sequencing a read could be thoroughly controlled in order to always deliver proper results.

### Challenge 2

In this example, two reads are already aligned. In general case, we could apply alignment, such as msa::msaClustalOmega, before calculate the transition/transversion ratio.

### Challenge 3

It is important to use gregexpr2 from Biostrings package, the other similar function (base::gregexpr) will only return non-overlap substrings.

### Challenge 4

- The complexity of my function to find all maximal repeat of a string is at least \$O(n\^2)\$ where \$n\$ is the length of the string. Thus, it can grow very quickly when the string is long.

- I believe that there is more optimal algorithm tackling that problem, such as deploying a suffix-tree algorithm when working with string.

- I set the threshold for number of maximal repeat to save time.

### Challenge 5

I used Fastqc on Bioinformatics server and download the results to PC for processing.

### Challenge 6

- The complexity of the first version algorithm depends on the number of combinations from the $\Delta(X)$. Thus, it could be extremely big when dealing with a big multiset. Furthermore, the algorithm will spend a lot of time checking the validity of those combinations.

- In the updated version, I add two more keys observations which allows to significantly reduce the running time:

  - Shifting all values in a set $X$ such that its smallest value to 0, this trick could save us computation of solving system of linear algebra as in the first version.

  - Let $X^{(0)}$ be the set of $X$ after shifting to 0. Then we have $X^{(0)} \cup \Delta(X^{(0)}) = L$. Thus, adding new element to find the original set $X$ must satisfy the property: $\Delta(X^{(0)'}) \subseteq \Delta(X) \setminus X^{(0)'}$.
